Relict deciduous oaks show high resilience to drought along a climatic gradient in the rear-edge despite strong land-use legacies 

### 2019-08-14 man/ecosystems/ms_ecosystems.Rmd
update new version

## Authors [^\*^] 
A.J. Perez-Luque^1,2,a^; G. Gea-Izquierdo^3,b^ and Zamora, R.^1,2,c^

@@

[^\*^]: Author Contributions. AJPL, GGI and RZ conceived the study, conducted field work and collected the data. AJPL and GGI conducted lab work and led article preparation. AJPL analyzed data and led the writing of the paper. GGI and RZ contributed in the writing process. All authors contributed to the drafts and gave final approval for publication.

@@

* The response to climate and resilience followed a water-stress gradient. Primary and secondary growth of more xeric populations (*i.e.* Northern aspect and low elevations) was less resilient to drought than that at more mesic sites. In addition, a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic high-elevation stands, but absent in the most xeric site. Trees were highly sensitivity to drought and both primary and secondary growth expressed high resilience to drought events in the short- and the long-term. The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood extraction) in the current forest structure.

@@

* Despite being considered relict stands at the dry rear-edge for the species, the trees showed high resilience in response to drought. Yet, there were a divergent response to a changing climate along very short climatic gradients. The oaks were more sensitive to water availability expressed at the most xeric sites whereas they exhibited positive growth trends only in the coldest and more humid stands. These results suggest that stand dynamics, particularly at the lowest elevations (*i.e.* the altitudinal rear-edge), need to be monitored under future expected drier and warmer climatic scenarios bearing in mind that management legacies likely play a key role in current stand dynamics.

@@

The Iberian Peninsula suffered several extreme drought episodes in the last decades of the twentieth century (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3). We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term (see Table S3). The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015. We used SPEI-12 months scale for all 0.5º grid cells covering Sierra Nevada obtained from Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts in the month when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values are below the threshold for at least two consecutive months. Several indicators were computed for each drought event: *duration* as the number of consecutive months with the SPEI lower than a certain threshold; *severity* of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.

@@

# Discussion
## Relict oaks show high resilience to recent drought events and long-term climatic variability

Severe drought negatively affects both primary and secondary growth of *Q. pyrenaica* forests. This was expressed by the observed reduction in greenness and tree-growth in response to the 2005 and 2012 drought events, and the consistent radial growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014]. Furthermore, the greatest reduction of tree-growth was observed during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018]. Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g.* 2005 and 2012 were considered two of the worst recorded drought; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* for the last 16 years (78.9 % of the pixels showed a positive trend during the 2000 to 2016 period). This confirms previous findings that pointed out a recent short-term increase in primary productivity for these forests coinciding with a rather humid decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi]. For tree-growth, we also observed positive trends in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along its distribution range only at high elevation sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold limited high elevation sites [@Salzer2009; @GeaIzquierdo2014]. However, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].

Despite the 2012 drought event was more severe and intense than 2005 (Table S3), resilience values for greenness and tree-growth were greater for 2012. This could be explained by the different timing of the two droughts, *i.e.* the 2012 drought was a winter drought [@Trigo2013] occurring earlier than the 2005 drought. The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q. pyrenaica* forests (Figure S7). Thus, this highlights the importance of the timing of the drought as a key factor determining tree recovery after drought [@Huang2018], as also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b]. In addition, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3), matching similar declining patterns observed for other species [@SanchezSalguero2012; @Camarero2015]. But, interestingly, the highest values of resilience for tree-growth were obtained for these two drought events. In addition, we observed a positive relation between the tree-growth resilience and drought severity (Figures 6). This result seems contradict our initial hypothesis in which we predicted low resilience values for this oak species in their rear-edge after extreme drought events. Yet, the high resilience values obtanied for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality (F.J. Navarro *personal communications*), which would show the influence of stand dynamics partly shadowing interpretation of individual resilience estimated from individual growth series.

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005]. A high vulnerability to drought is often assumed for populations located at their rear-edge [@MartinezVilalta2018]. However we obtained positive trends in greenness and tree-growth for the rear-edge of *Q. pyrenaica* and high values of resilience to severe droughts. Our findings are in agreement with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017]. @MartinezVilalta2018 pointed out the importance of the local adaptation and plasticity, and also of the local environmental factors on the vulnerability showed by rear-edge populations. The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada [@Valbuena2013], seem to be indicative of high local adaptation of this oak. In addition the local environmental conditions of the sites where this species inhabit at Sierra Nevada would explain the low vulnerability to drought showed by this oak. A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@MartinezVilalta2018]. In this sense, the high values of resilience to the drought events that we have observed, could suggest that the populations of *Q. pyrenaica* in Sierra Nevada are located in a geographical but not a climatic rear-edge. And the current niche is a result of land-use changes, which further complicates the definition of potential rear-edges.

Greenness and tree-growth of *Q. pyrenaica* showed different sensitivity to severe droughts. During the last drought events EVI was reduced to 90.3 % (Table 1), whereas BAI was reduced to 70.8 % (Table S2), suggesting a lower sensitivity to drought of greenness than of tree-growth, particularly for drier sites. These findings are in accordance with previously works showing that tree-growth is a more sensitive metric of forest resilience than net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018]. Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based (*i.e.* ecosystem or stand based); the growth reduction seems to be more mediated by sink that by source limitations [@Gazol2018]. The response of greenness to drought not only reflects the response of trees but from all vegetation within the pixel [@Dionisioetal2012IPoGEC]. In this sense, recent results indicate that remote sensing models underestimate the impact of drought on primary production [@Stocker2019], as soil moisture is not taken into account by this satellite estimates. This is particularly important for *Q. pyrenaica* forests, as soil moisture is a key growth-related factor for this species [@GeaIzquierdo2014]. On the other hand, tree-ring records allow to complement remote sensing data in a longer time scale, by reflecting climate or disturbance induced tree-growth anomalies over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018]. A combined analysis, such the approach used in our study, may allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes [@Babst2017], since spatio temporal patterns of tree-ring measurement and earth observations (*i.e.* EVI) reflects processes at different spatio-temporal scales [@GeaIzquierdo2015; @Babst2018].

@@

It has been described an effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive, thus high-elevations can be used as local refuge for species in response to climate change [@Bhuyan2017]. It is well known that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM]. Remarkably, the trees at CA-High, which are located around 1900 m. *a.s.l.* and represent today the upper altitudinal limit (treeline) for the species in this southernmost location, showed higher BAI than those located at lower-elevations (CA-Low and SJ) (Figure 4). This shows the high variability in the response to climate exhibited in a small gradient and is especially interesting for the southern sites, which are very close to each other, and overall both constitute the rear-edge for the species. The higher values of tree growth in taller trees that we observed in CA-High site strongly suggest a lower water stress at high-elevation sites [as shown by soil analyses along an elevational gradient in this location @CoboDiaz2017]. Furthermore, as mentioned, the recent growth increase only at the most humid locations a likely concomitant positive effect of a warming climate [@GeaIzquierdo2014].

Finally and in a context of global change, in addition to the withstanding lower water stress, it is likely that the less accessible sites at higher elevations would suffer a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources (Table S4).

@@

## Land-use legacies shape forests and sensitivity to climate change at the rear-edge

The Mediterranean landscapes, even at high elevations, have suffered a strong transformation driven by human activities throughout history. Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@GarciaAlix2017]. From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with a sharp intensification of the human activities in the last 150 years [@GarciaAlix2017]. Mining, forest clearings, firewood extraction and charcoal exploitations (these were strongly suggested in the disturbance chronology in SJ), pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada, and on Pyrenean oak forest, as can be derived from the results of the review of historical documents. As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia2015]. Therefore, our study sites have been subjected to an intense exploitation of forest resources in history and landscape transformation. Although the estimated age for our sampled trees was lower than 180 years, several documents reported abundant oak presence before that date. For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, reported for Cáñar and SJ sites more than two millions and 700 000 trees respectivley, most of them juveniles, suggesting recent wood fellings [@Cruz1991]. These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g. @GeaIzquierdo2014; @Gea-Izquierdo2015; @Dorado2017].

Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between Northern (coppice) and Southern populations (high forest, open woodland) [@JimenezOlivencia2015]. It has been described a sequentially distribution of the land uses along the elevational gradient in the northern site (*e.g.* San Juan): grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops [@JimenezOlivencia2015]. Other activities (like mining) should have determined the forest structure at that northern location, sincr SJ is located in an area with a high concentration of mines and quarries that have been exploited intermitently throughout history. On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands along the elevational gradient where oaks occurr [@Calatrava2019], being firewood, charcoal and acorns continuously exploited through history [@Valbuena2013]. Charcoal extraction is weel known at least since the 15th century up to middle of the 20th century [@JimenezSerrano2004]. From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment and use of gas and fosil fuels, similar to other Fagaceae stands across Europe and the Mediterranean [@Valbuena2013].

Some of the previous historical records of land-use were expressed in our growth chronologies. The release event at 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral [@Titos1990]. This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth the remnant trees included in the northern site (Figure S4). In addition, paleoecological studies carried out in nearby alpine bogs have recorded increases in the heavy metals concentration since the end of 18th century until mid-20th, which coincides with the maximum activity of the mining in the northern site [@GarciaAlix2017]. The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 5, S4). There are no records of forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest] and no record for any logging  during 1995 - 2000 period (F.J. Cano-Manuel *personal communications*). Therefore, two non-mutually exclusive ways could explain the tree-growth release observed. Firstly it could be related with a natural drought-induced mortality event after 1995 as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004]. On the other hand, the strong positive correlations of SPEI with tree-growth for this site show a high sensitivity to water availability [@GeaIzquierdo2014]. Previous works reported a sharp decline (or non-production) in latewood production during extreme drought events [@Corcuera2006]. Since latewood are less vulnerable to embolism than earlywood vessels [@Corcuera2006], the reduction or even the non-production of former could negatively affect to tree-growth and also enhanced the mortality, particularly for drier sites [@Corcuera2006; @GeaIzquierdo2014].

For southern sites, the only positive release event showed at the earliest yers for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] documented in this site [@Valbuena2013].

... revisar 
Although the abandonment of the traditional activities have provoked a stagnation for this species [@Canellas2004], we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource. This pattern concurs with a forest expansion into marginal abandoned croplands and an increase in the tree-cover [@CamachoOlmedo2002].  
...

>>>>

## Authors [^\*^] 
A.J. Pérez-Luque^1,2,a^; G. Gea-Izquierdo^3,b^ and Zamora, R.^1,2,c^

@@

[^\*^]: Author Contributions. AJPL, GGI and RZ conceived the study, conducted field work and collected the data. AJPL and GGI conducted lab work. AJPL analyzed data and led the writing of the paper. GGI and RZ contributed in the writing process. All authors contributed to the drafts and gave final approval for publication.

@@

* The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood extraction) in the current forest structure. Trees were highly sensitivity to drought and both primary and secondary growth expressed high resilience to drought events in the short- and the long-term. The response to climate and resilience followed a water-stress gradient. Primary and secondary growth of more xeric populations (*i.e.* northern aspect) was less resilient to drought than that at more mesic sites. In addition, a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic high-elevation stands, but absent in the most xeric site.

@@

@@

The Iberian Peninsula suffered several extreme drought episodes in the last decades of the twentieth century (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3). We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term (see Table S3). The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015. We used SPEI-12 months scale for all 0.5º grid cells covering Sierra Nevada obtained from Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts in the month when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values are below the threshold for at least two consecutive months. Several indicators were computed for each drought event: *duration* as the number of consecutive months with the SPEI lower than a certain threshold; *severity* of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.

@@

# Discussion
## Land-use legacies shape forests and sensitivity to climate change at the rear-edge

The Mediterranean landscapes, even at high elevations, have suffered a strong transformation driven by human activities throughout history [@GarciaAlix2017]. Forest clearings, firewood extraction, charcoal exploitations and mining (these were strongly suggested in the disturbance chronology in SJ), have strongly impacted on the forest resources at Sierra Nevada as can be derived from the results of the review of historical documents (Table S4). Indeed a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation has been documented in this mountain region [@JimenezOlivencia2015]. 
Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland) [@JimenezOlivencia2015]. It has been described a sequentially distribution of the land uses along the elevational gradient in the northern slopes of Sierra Nevada (*e.g.* SJ site): grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops [@JimenezOlivencia2015]. Other activities (like mining) should have determined the forest structure at that northern location, since SJ is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history. On the other hand, woodland areas of the southern slopes (*e.g.* CA site) are mixed with a greater percentage of croplands along the elevational gradient where oaks occur [@JimenezOlivencia2015]. Firewood, charcoal extraction and acorns have been resources intensively exploited in southern sites, until at least the middle of the 20th century when these activities sharply decreased mainly due to rural abandonment and use of gas and fossil fuels [@Valbuena2013; @JimenezOlivencia2015].

Some of the previous historical records of land-use were expressed in our growth chronologies. The release event at 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral [@Titos1990]. This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth the remnant trees included in the northern site (Figure S4). The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 5, S4). There are no records of forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest] and no record for any logging  during 1995 - 2000 period (F.J. Cano-Manuel *personal communications*). We speculate that this might be related with a natural drought-induced mortality event after 1995 as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004]. On the other hand, the strong positive correlations of SPEI with tree-growth for this site show a high sensitivity to water availability [@GeaIzquierdo2014]. For southern sites, the only positive release event showed at the earliest years for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] documented in this site [@Valbuena2013].

As we inferred from our results, the management history can conditioned tree-growth and resilience of tree species, highlighting the importance of land-use legacies in Mediterranean forests [@Navarro2013; @DoblasMiranda2017]. Nevertheless, we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource. This pattern concurs with a forest expansion into marginal abandoned croplands and an increase in the tree-cover [@CamachoOlmedo2002].

```{r eval=FALSE, echo=FALSE}
~~Although the estimated age for our sampled trees was lower than 180 years, several documents reported abundant oak presence before that date. For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, reported for Cáñar and SJ sites more than two millions and 700 000 trees respectivley, most of them juveniles, suggesting recent wood fellings [@Cruz1991]. These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g. @GeaIzquierdo2014; @GeaIzquierdo2015EJFR; @Dorado2017].~~  
```

## Relict oaks show high resilience to recent drought events and long-term climatic variability

Severe drought negatively affects both primary and secondary growth of *Q. pyrenaica* forests. This was expressed by the observed reduction in greenness and tree-growth in response to the 2005 and 2012 drought events, and the consistent radial growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014]. Furthermore, the greatest reduction of tree-growth was observed during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018]. Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g.* 2005 and 2012 were considered two of the worst recorded drought; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* for the last 16 years, which is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather humid decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto]. For tree-growth, we also observed positive trends in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along its distribution range only at high elevation sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold limited high elevation sites [@Salzer2009; @GeaIzquierdo2014]. However, the observed pattern for our sites (a rear-edge) differs from the declining trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].

Despite the 2012 drought event was more severe and intense than 2005 (Table S3), resilience values for greenness and tree-growth were greater for 2012. This could be explained by the different timing of the two droughts, *i.e.* the 2012 drought was a winter drought [@Trigo2013] occurring earlier than the 2005 drought. The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q. pyrenaica* forests (Figure S7). Thus, this highlights the importance of the timing of the drought as a key factor determining tree recovery after drought [@Huang2018], as also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b]. For tree-growth, the highest values of resilience were obtained for the most severe events, namely 1995 and 1999 (Table S3). In addition, we observed a positive relation between the tree-growth resilience and drought severity (Figures 6). This result seems to contradict our initial hypothesis in which we predicted low resilience values for this oak species in their rear-edge after extreme drought events. Yet, the high resilience values obtained for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality (F.J. Navarro *personal communications*), which would show the influence of stand dynamics partly shadowing interpretation of individual resilience estimated from individual growth series.

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005]. A high vulnerability to drought is often assumed for populations located at their rear-edge [@MartinezVilalta2018]. However we obtained positive trends in greenness and tree-growth for the rear-edge of *Q. pyrenaica* and high values of resilience to severe droughts. Our findings are in agreement with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017]. @MartinezVilalta2018 pointed out the importance of the local adaptation and plasticity, and also of the local environmental factors on the vulnerability showed by rear-edge populations. The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada [@Valbuena2013], seem to be indicative of high local adaptation of this oak. In addition the local environmental conditions of the sites where this species inhabit at Sierra Nevada would explain the low vulnerability to drought showed by this oak. A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@MartinezVilalta2018]. In this sense, the high values of resilience to the drought events that we have observed, could suggest that the populations of *Q. pyrenaica* in Sierra Nevada are located in a geographical but not a climatic rear-edge. And the current niche is a result of land-use changes, which further complicates the definition of potential rear-edges. Similarly, high values of resilience showed by other Mediterranean species located in their rear-edge [*e.g.* Pinus sylvestris, @Herrero2014] seem to be related with stabilizing mechanisms promoting community resilience that can buffer the impacts of extreme events.

Greenness and tree-growth of *Q. pyrenaica* showed different sensitivity to severe droughts. During the last drought events EVI was reduced to 90.3 % (Table 1), whereas BAI was reduced to 70.8 % (Table S2), suggesting a lower sensitivity to drought of greenness than of tree-growth, particularly for drier sites. These findings are in accordance with previously works showing that tree-growth is a more sensitive metric of forest resilience than net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018]. Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based (*i.e.* ecosystem or stand based); the growth reduction seems to be more mediated by sink that by source limitations [@Korner2013]. The response of greenness to drought not only reflects the response of trees but from all vegetation within the pixel [@Dionisioetal2012IPoGEC]. In this sense, recent results indicate that remote sensing models underestimate the impact of drought on primary production [@Stocker2019], as soil moisture is not taken into account by this satellite estimates. This is particularly important for *Q. pyrenaica* forests, as soil moisture is a key growth-related factor for this species [@GeaIzquierdo2014]. On the other hand, tree-ring records allow to complement remote sensing data in a longer time scale, by reflecting climate or disturbance induced tree-growth anomalies over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018]. A combined analysis, such the approach used in our study, may allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes [@Babst2017], since spatio temporal patterns of tree-ring measurement and earth observations (*i.e.* EVI) reflects processes at different spatio-temporal scales [@GeaIzquierdo2015; @Babst2018].

@@

It has been described an effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive, thus high-elevations can be used as local refuge for species in response to climate change [@Bhuyan2017]. It is well known that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM]. Remarkably, the trees at CA-High, which are located around 1900 m. *a.s.l.* and represent today the upper altitudinal limit (treeline) for the species in this southernmost location, showed higher BAI than those located at lower-elevations (CA-Low and SJ) (Figure 4). This shows the high variability in the response to climate exhibited in a small gradient and is especially interesting for the southern sites, which are very close to each other, and overall both constitute the rear-edge for the species. The higher values of tree growth in taller trees that we observed in CA-High site strongly suggest a lower water stress at high-elevation sites [as shown by soil analyses along an elevational gradient in this location @CoboDiaz2017]. Furthermore, as mentioned, the recent growth increase only at the most humid locations a likely concomitant positive effect of a warming climate [@GeaIzquierdo2014]. Finally and in a context of global change, in addition to the withstanding lower water stress, it is likely that the less accessible sites at higher elevations would suffer a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources (Table S4).

@@
